---
title: "Transposable elements (TE) analysis in public datasets spanning pre- and postimplantation mouse development"
date: "April 11th, 2023"
author: 
- name: Sara Formichetti
  affiliation: EMBL Rome
  email: sara.formichetti@embl.it
output:
  prettydoc::html_pretty:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---

```{r Setting general options, include=FALSE}

knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')

```

```{r loading needed packages}

library(ggplot2)
library(data.table)
library(pheatmap)
library(DESeq2)
library(gridExtra)
library(ggrepel)
library(ggpubr)

```

# Mapping and TE counts stats

Note that:

* STAR parameters were adjusted for trasposable elements analysis (and can be found in embryo_public_data_GSE66582_GSE76505_reanalysis/snake-make/TE_RNASeq_w_Clontech_trim.Snakefile)
* featureCounts was run using RepeatMasker annotation and for both all TE and only full-length (fl) ones

```{r load sequencing stats}

seq_stat_files <- c("STAR_input_reads.txt", "featureCounts_assigned.txt", "featureCounts_assigned_fl.txt", "STAR_uniquely_mapped.txt", "STAR_multiple_loci.txt", "STAR_too_many_loci.txt")
names(seq_stat_files) <- gsub("\\.txt", "", seq_stat_files)
STAR_TE_folder <- "../../data/sequencing/stats/STAR_TE/"

```

```{r samples table}

# loading samples table
samples_table <- fread("../../data/sequencing/samples_table_TE_analysis.csv")
names(samples_table)[grep("group_or_time_point", names(samples_table))] <- "stage"

```

```{r STAR stats per stage, include = TRUE}

get_stat_list <- function (my_stat_folder, i, my_stat_files) {
  # by using fread w these sep and sep2 parameters, the numeric column is always V2
  stat_dt <- fread(paste0(my_stat_folder, my_stat_files[i]), sep = "\t", sep2 = " ")
  names(stat_dt) <- c("sample", names(my_stat_files)[i])
  stat_dt$sample <- gsub("_TE_counts.*", "", gsub("_amgm350.*", "", gsub(".*/", "", stat_dt$sample)))
  return(stat_dt)
}

stat_list <- lapply(1:length(seq_stat_files), get_stat_list, my_stat_folder = STAR_TE_folder, my_stat_files = seq_stat_files)
all_stat_dt <- Reduce(function(...) merge(..., by = c("sample")), stat_list)
my_cols <- c("sample", names(seq_stat_files)[-1])
all_stat_melt_dt <- melt(all_stat_dt[, ..my_cols], variable.name = "stat")
my_cols <- c("sample", "STAR_input_reads")
all_stat_melt_dt<- merge(all_stat_melt_dt, all_stat_dt[, ..my_cols], by = "sample")
stats_sample_table <- merge(all_stat_melt_dt, samples_table, by = "sample")
stats_sample_table$stage <- factor(stats_sample_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))

ggplot(data = stats_sample_table, aes(x = stage, y = value/STAR_input_reads*100)) +
  geom_point() +
  facet_wrap(vars(stat), ncol = 2, scales = "free_y") +
  xlab("") +
  ylab("% of reads") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r rDNA perc per stage, include = TRUE}

# reading rDNA stats
rDNA <- fread("../../data/sequencing/stats/rDNA_stats.txt")
names(rDNA) <- c("sample", "rDNA_perc")
rDNA_sample_table <- merge(rDNA, samples_table, by = "sample")
rDNA_sample_table <- merge(rDNA_sample_table, all_stat_dt[,.(sample,STAR_input_reads)], by = "sample")
rDNA_sample_table_m <- melt(rDNA_sample_table, measure.vars = c("rDNA_perc", "STAR_input_reads"))
rDNA_sample_table_m$stage <- factor(rDNA_sample_table_m$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))

ggplot(data = rDNA_sample_table_m, aes(x = stage, y = value)) +
  geom_point(aes(color = stage), show.legend = FALSE) +
  geom_text_repel(aes(label = sample), size = 2, max.overlaps = 100) +
  xlab("") +
  facet_wrap(facets = vars(variable), ncol = 2, scale = "free_y") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

# Reading TE counts

```{r load featureCounts output tables}

# specifying the directory containing featureCounts outputs
dir <- "../../data/sequencing/featureCounts/STAR_TE"

# getting file names of featureCounts output count tables, sorted alphabetically
files <- sort(list.files(path = dir)[grep("counts.txt$", list.files(path = dir))])

# character vector of names of files with path
files_w_path <- file.path(dir, files)

# reading the count tables: first line is a commented line containing featureCounts command and needs to be skipped; columns chr,start,end,strand are dropped because they contain the comma-separated list of chr,start,end,strand for all occurrences of the TE with that unique repName in the annotation - in case of single copy genes these columns usually contain one value, in this case a list of them because each element is repeated in the genome.
counts_list <- lapply(files_w_path, function (f) {fread(f, skip = 1, drop = c("Chr", "Start",	"End", "Strand"))})
# all counts tables have the same dimensions because featureCounts counts for all meta-features (and puts 0 in case there are 0 count for that feature in that sample), therefore there will be no NAs in the reduced dt
counts_dt <- Reduce(function(...) merge(..., by = c("Geneid", "Length"), all = TRUE), counts_list)
# remove path from sample names
names(counts_dt) <- sub(".*/", "", sub("_amgm350.TE_only.bam", "", names(counts_dt)))
names(counts_dt)[1] <- "repName"

rm(counts_list)

```

```{r merging counts with TE annotation}

# loading TE annotation
TE_annotation_bed <- "../../data/annotations/RepeatMasker_RepeatLibrary20140131_mm10.noGenes.noSimple.bed"
TE_anno <- fread(TE_annotation_bed, fill = TRUE, col.names = c("seqnames", "start", "end", "repName", "strand", "repFamClass"), drop = 5)
# adding +1 to start is necessary for compatibility with methylBase-derived objects because bed file coordinates are 0-based
TE_anno$start <- TE_anno$start + 1
# making column with repFamilies only and column with repClasses only
TE_anno$repFamily <- gsub(".*/", "", TE_anno$repFamClass)
TE_anno$repClass <- gsub("/.*", "", TE_anno$repFamClass)

# merging the count table with TE annotation to get repFamilies and Classes
counts_dt <- merge(counts_dt, unique(TE_anno[,c("repName", "repClass", "repFamily", "repFamClass")]), by = "repName", all.x = TRUE, all.y = FALSE)
# removing rows whose repFamClass is NA
counts_dt <- counts_dt[!is.na(counts_dt$repFamClass)]

```

The input files are:

* the output tables made by featureCounts (run in rule 'count_on_TE' of the Snakefile), which contain read counts of all repetitive elements (i.e. all repNames) for all samples
* the repetitive elements annotation `r TE_annotation_bed` (converted to SAF format in rule of Snakefile) used for featureCounts command, that is used to retrieve repFamily and repClass for each repName. The annotation I chose is the most updated library to Nov2020 of rmsk, from which I removed 'Simple repeats' and 'Low complexity regions'
* a table with total number of reads per sample in raw fastq files 
* the samples' table, containing metadata (i.e. stage)

```{r ordering based on stage}

# releveling stage for plotting
samples_table$stage <- factor(samples_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))
setkey(samples_table, stage)
# ordering counts dt based on samples table
setcolorder(counts_dt, c("repName", "repClass", "repFamily", "repFamClass", samples_table$sample))

```

# Analysis at family level 

## FPKM for each family of Repetitive Elements

For each family of Repetitive Elements (in case of elements with no repFamily name or repFamilies belonging to more than one repClass I use repClass) I compute **FPKM** values, as follows: for each sample:

* I compute the sum of counts for all elements belonging to that repFamily
* I divide this sum by the total number of reads for that sample and multiply by 10â¶
* I divide this number by the total sum of lengths (in Kb) of the elements belonging to that repFamily --> FPKM
* When specified, **I subtract from each FPKM the total FPKM of all transposons belonging to the DNA repClass**

```{r load library size table}

# loading table containing total number of STAR input reads per sample
library_size_dt <- fread("../../data/sequencing/stats/STAR_TE/STAR_input_reads.txt")
library_size_dt$sample <- gsub(".*/", "", gsub("_amgm350.*", "", library_size_dt$V1))
library_size_dt$STAR_input <- library_size_dt$V2
library_size_dt <- library_size_dt[, .(sample, STAR_input)]
# since I counted fragments in featureCounts, I divide the STAR input reads by 2 to get a rough estimation of read pairs
library_size_dt$library_size <- library_size_dt$STAR_input/2
# removing bad and poorly injected samples
library_size_dt <- library_size_dt[library_size_dt$sample %in% samples_table$sample, ]
# ordering library size as samples and counts table
library_size_dt <- library_size_dt[match(samples_table$sample, library_size_dt$sample)]

```

```{r family FPKM}

# I notice that some elements do not have a repFamily but they have a repClass and that some repFamilies are assigned to more than one repClass, therefore I use 'repFamClass' to compute the FPKM 
# selecting columns on which to compute the sum by 'repFamClass'
cols <- c("Length", names(counts_dt)[names(counts_dt) %in% samples_table$sample])
# sum by 'repFamClass'
family_counts_dt <- counts_dt[, lapply(.SD, sum), by = .(repFamClass), .SDcols = cols] 
# remove rows containing only 0
family_counts_dt <- family_counts_dt[rowSums(family_counts_dt[, cols[2:length(cols)], with = FALSE]) > 0, ]
# divide each column by library size and multiply by 10^6
family_FPKM_dt <- sweep(family_counts_dt[, cols[2:length(cols)], with = FALSE], 2, library_size_dt$library_size/1000000, FUN = '/')
# divide each row by repFamClass total length
family_FPKM_dt <- as.data.table(sweep(family_FPKM_dt, 1, family_counts_dt$Length/1000, FUN = '/'))
family_FPKM_dt$repFamClass <- family_counts_dt$repFamClass
# subtract DNA transposons' FPKM from each other family's FPKM value
DNA_TE_FPKM <- setNames(as.numeric(colSums(family_FPKM_dt[grep("DNA/", family_FPKM_dt$repFamClass), 1:(ncol(family_FPKM_dt)-1)])), cols[2:length(cols)])
family_FPKM_minusDNA_dt <- as.data.table(sweep(family_FPKM_dt[, 1:(ncol(family_FPKM_dt)-1)], 2, DNA_TE_FPKM, FUN = "-"))

# add back columns with repFamilies and repClasses
family_FPKM_dt$repFamily <- gsub(".*/", "", family_FPKM_dt$repFamClass)
family_FPKM_dt$repClass <- gsub("/.*", "", family_FPKM_dt$repFamClass)
family_FPKM_minusDNA_dt$repFamClass <- family_FPKM_dt$repFamClass
family_FPKM_minusDNA_dt$repFamily <- family_FPKM_dt$repFamily
family_FPKM_minusDNA_dt$repClass <- family_FPKM_dt$repClass

```

## Heatmaps

The heatmaps are scaled by rows.

```{r heatmap function}

# Custom function for heatmap of repetitive elements from data.table of FPKM
all_samples_TE_heatmap <- function (my_measures_table, my_samples_table, sample_cluster, elements_cluster, my_title, scale_col) {
  # making df with only the numeric values
  pheat_df <- as.data.frame(my_measures_table)[, colnames(my_measures_table) %in% my_samples_table$sample]
  # rownames appearing on heatmap will be the repFamily names (which are actually the repClass when there is no repFamily assigned to an element)
  repFamilies <- gsub(".*/", "", my_measures_table$repFamClass)
  # in cases when one repFamily name is associated to more than one repClass, duplicates are created in repFamilies vector. Using repFamClass in these cases
  ambigous_rows <- which(duplicated(repFamilies) | duplicated(repFamilies, fromLast=TRUE))
  repFamilies[ambigous_rows] <- my_measures_table$repFamClass[ambigous_rows]
  rownames(pheat_df) <- repFamilies
  # ordering stages
  my_samples_table$stage <- factor(my_samples_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))
  setkey(my_samples_table, stage)
  pheat_df <- pheat_df[, my_samples_table$sample]
  # annotating samples with metadata
  pheat_anno_samples <- data.frame(stage = my_samples_table$stage, row.names = my_samples_table$sample)
  # annotating elements with repClasses
  pheat_anno_elem <- data.frame(repClass = my_measures_table$repClass, repFamilies, row.names = rownames(pheat_df))
  # ordering TE for better visualization in case no elements clustering is required
  pheat_df <- pheat_df[order(pheat_anno_elem$repClass), ]
  pheat_anno_elem <- pheat_anno_elem[rownames(pheat_df), ]
  pheat_anno_elem$repFamilies <- NULL
  # removing rows and columns having only 0s
  pheat_df <- pheat_df[rowSums(pheat_df) != 0, colSums(pheat_df) != 0]
  if (scale_col) {
    rn <- rownames(pheat_df)
    pheat_df <- as.data.frame(apply(pheat_df, 2, scale))
    rownames(pheat_df) <- rn
  }
  # plotting heatmap
  pheatmap(pheat_df, cluster_rows = elements_cluster, cluster_cols = sample_cluster, annotation_col = pheat_anno_samples, annotation_row = pheat_anno_elem, show_rownames = TRUE, clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", fontsize = 12, scale = "row", main = my_title)
}

```

### Preimplantation stages only

```{r heatmap subset of elements preimpl, include = TRUE, fig.height=12, fig.width=15}

preimpl_samples_table <- samples_table[samples_table$batch == 1,]

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

### Postimplantation stages only

```{r heatmap subset of elements postimpl, include = TRUE, fig.height=12, fig.width=15}

postimpl_samples_table <- samples_table[samples_table$batch == 2,]

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = postimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = postimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = postimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = postimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

### All stages

```{r heatmap subset of elements all, include = TRUE, fig.height=12, fig.width=15}

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

Based on all above plots, I remove the E5.5 VE samples and early 2-cell sample SRR2927028 and replot.

```{r removing bad samples}

# removing E5.5 samplesbased on rDNA percentage
samples_table <- samples_table[!(samples_table$sample %in% c("SRR2927028", "SRR3083899", "SRR3083907", "SRR3083908", "SRR3083903_SRR3083904", "SRR3083905_SRR3083906"))]
# releveling stage for plotting
samples_table$stage <- factor(samples_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))
setkey(samples_table, stage)
# ordering counts dt based on samples table
setcolorder(counts_dt, c("repName", "repClass", "repFamily", "repFamClass", samples_table$sample))

```

```{r heatmap subset of elements all after filtering, include = TRUE, fig.height=12, fig.width=15}

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

Note that when correcting for DNA FPKM preimplantation embryonic stages get penalized compared to the oocyte i.e. oocyte FPKM values are overestimated.

# DE-Seq analysis of RNA transposons - preimplantation stages only

I include the **FPKM of DNA repetitive elements as confounding factors in DESeq2 formula**.

```{r creating DESeq inputs}

preimpl_samples_table <- samples_table[samples_table$batch == 1,]
preimpl_counts_dt <- counts_dt[, c("repName", "repClass", "repFamily", "repFamClass", preimpl_samples_table$sample, "Length"), with = FALSE]

# Selecting only RNA transposons and only columns with numeric values
RNA_TE_counts_dt <- preimpl_counts_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", preimpl_counts_dt$repClass), preimpl_samples_table$sample, with = FALSE]

# A count matrix will be the input for DESeq
counts_m <- as.matrix(RNA_TE_counts_dt)
rownames(counts_m) <- preimpl_counts_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", counts_dt$repClass), ]$repName
# the design df is created from the samples_table
design_df <- data.frame(stage = preimpl_samples_table$stage, batch = preimpl_samples_table$batch)
# the rownames of the design df have to be the same (and in the same order) of the colnames of the count matrix
rownames(design_df) <- preimpl_samples_table$sample
# adding another DNA transposons FPKM as an additional variable, that can be used as counfounding factor
design_df$sample <- rownames(design_df)
design_df$DNA_FPKM <- DNA_TE_FPKM[design_df$sample]
design_df$batch <- as.character(design_df$batch)

```

```{r creating DESeq Dataset}

dds <- DESeqDataSetFromMatrix(countData = counts_m,
                              colData = design_df,
                              design = ~ DNA_FPKM + stage)

```

Before running the Differential Expression analysis, the data are pre-filtered to remove all repetitive elements with < 10 reads among all samples.

```{r prefiltering}

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

```

```{r DESeq function}

dds <- DESeq(dds)

```

# Results

```{r adj pvalue cutoff}

adjpvalue_cutoff <- 0.1
l2fc_cutoff_to_label <- 0.2
baseMean_cutoff_to_label <- 10

```

```{r DESeq results list}

# Function which extracts DESeq results (with shrunken log2FC) for all comparisons
get_results_all_comp <- function (my_dds) {
  res_list <- list(
    zygote = lfcShrink(my_dds, contrast=c("stage","zygote","MII_oocyte"), alpha=adjpvalue_cutoff, type="ashr"),
    preEGA_2c = lfcShrink(my_dds, contrast=c("stage","early_2cell","zygote"), alpha=adjpvalue_cutoff, type="ashr"),
    postEGA_2c = lfcShrink(my_dds, contrast=c("stage","2cell","early_2cell"), alpha=adjpvalue_cutoff, type="ashr"),
    postEGA_2c_to_zygote = lfcShrink(my_dds, contrast=c("stage","2cell","zygote"), alpha=adjpvalue_cutoff, type="ashr"),
    fourCell = lfcShrink(my_dds, contrast=c("stage","4cell","2cell"), alpha=adjpvalue_cutoff, type="ashr"),
    eightCell = lfcShrink(my_dds, contrast=c("stage","8cell","4cell"), alpha=adjpvalue_cutoff, type="ashr"),
    ICM = lfcShrink(my_dds, contrast=c("stage","Inner_cell_mass","8cell"), alpha=adjpvalue_cutoff, type="ashr"))
  return(res_list)
}

# extracting result table from the DESeq analysis for all comparisons
res_list <- get_results_all_comp(dds)

```

```{r ranked DE genes}

# Function which takes a DESeq results object and return it as a dataframe, with genes ordered by pvalue and filtered by adjusted pvalue
pvalue_ranked_DE <- function (my_res) {
  # keeping only genes with adjusted pvalue below my threshold
  my_res_f <- subset(my_res, padj < adjpvalue_cutoff)
  # ordering by pvalue
  my_res_f_ord <- my_res_f[order(my_res_f$pvalue),]
  my_res_f_ord$repName <- rownames(my_res_f_ord)
  return(my_res_f_ord)
}

pvalue_ranked_DE_list <- lapply(res_list, pvalue_ranked_DE)

```

I exclude the comparisons including the preEGA 2c stage because, after filtering bad samples, this stage includes only one biological replicate. 

```{r DE TEs, results='asis'}

knitr::kable(as.data.frame(pvalue_ranked_DE_list[[1]]), caption = "zygote vs MII oocyte")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[4]]), caption = "postEGA 2c vs zygote")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[5]]), caption = "fourCell vs postEGA 2c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[6]]), caption = "eightCell vs postEGA 2c")

```

## MA-plots

```{r customizing plots DESeq2}

pval_hist <- function (my_dds_res) {
  ggplot(as(my_dds_res, "data.frame"), aes(x = pvalue)) +
  geom_histogram(binwidth = 0.01, fill = "Royalblue", boundary = 0) +
  ggtitle(gsub(".*: stage ", "", mcols(my_dds_res)$description[4])) +
  scale_x_continuous(breaks=c(0.0, 0.5, 1)) +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        title = element_text(size = 18),
        legend.text = element_text(size = 20), legend.title = element_text(size = 20),
        legend.key.height = unit(1.5,"line"))
}

MA_plot_TE <- function(my_dds_res, adjpval_to_color, l2fc_to_label, baseMean_to_label, ylims) {
  # adding label to genes which are significant and expressed above defined thresholds
  my_dds_res$to_label <- ""
  my_dds_res$to_label[my_dds_res$baseMean >= baseMean_to_label & my_dds_res$padj < adjpval_to_color & (my_dds_res$log2FoldChange >= l2fc_to_label | my_dds_res$log2FoldChange <= -l2fc_to_label)] <- rownames(my_dds_res[my_dds_res$baseMean >= baseMean_to_label & my_dds_res$padj < adjpval_to_color & (my_dds_res$log2FoldChange >= l2fc_to_label | my_dds_res$log2FoldChange <= -l2fc_to_label), ])
  my_dds_res$to_color <- ""
  my_dds_res$to_color[my_dds_res$padj < adjpval_to_color] <- paste0("adjpval < ", adjpval_to_color)
  dt_to_plot <- as.data.table(my_dds_res)
  dt_to_plot$repName <- rownames(my_dds_res)
  # adding color by repFamily to genes which are significant and expressed above defined thresholds
  dt_to_plot <- merge(dt_to_plot, unique(TE_anno[,.(repName,repFamily)]), by = "repName")
  dt_to_plot$to_color <- dt_to_plot$repFamily
  dt_to_plot$to_color[dt_to_plot$padj >= adjpval_to_color] <- ""
  my.cols <- brewer.pal(length(unique(dt_to_plot$to_color)), "Paired")
  my.cols[1] <- "grey"
  ggplot(dt_to_plot, aes(x = baseMean, y = log2FoldChange)) +
    geom_point(aes(color = to_color), alpha = 0.8) +
    geom_text_repel(aes(label = to_label), size = 4, max.overlaps = 10, show.legend = FALSE) +
    scale_x_log10() +
    ylim(ylims) +
    scale_color_manual(values = my.cols) +
    ggtitle(gsub(".*: stage ", "", mcols(my_dds_res)$description[4])) +
    theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
          title = element_text(size = 20),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black"),
          panel.border = element_rect(colour = "black", fill=NA))
}

```

In the MA-plots:

* The threshold used for a dot to be coloured based on repFamily is p-value adjusted < `r adjpvalue_cutoff`.
* Transposable elements whose mean expression > `r baseMean_cutoff_to_label` and log2FoldChange > `r l2fc_cutoff_to_label` (or < -`r l2fc_cutoff_to_label`) are labeled (but labeling is limited by max.overlap parameter).

```{r MA plots, include = TRUE, fig.width=27, fig.height=24}

library(RColorBrewer)
do.call(grid.arrange, lapply(res_list[c(1,4,5,6,7)], function (r) {MA_plot_TE(r, adjpval_to_color = adjpvalue_cutoff, l2fc_to_label = l2fc_cutoff_to_label, baseMean_to_label = baseMean_cutoff_to_label, ylims = c(-10,10))}))

```

```{r sessioninfo, setup, echo = TRUE, results = 'markup'}

sessionInfo()

```