---
title: "Comparison of morulae and blastocysts GSEA results"
date: '2022-09-16'
author: 
- name: Sara Formichetti
  affiliation: EMBL Rome
  email: sara.formichetti@embl.it
output:
  prettydoc::html_pretty:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---

```{r Setting general options, include=FALSE}

knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, echo = FALSE, warning = FALSE, message = FALSE, results = 'hide', fig.width=12, fig.height=8)

```

```{r loading needed packages}

library("DESeq2")
library(ggplot2)
library("ggrepel")
library("vsn")
library(data.table)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(DOSE)

```

```{r loading necessary Rdata}

load("Rdata/all_gsea_activeVsCtrl.Rdata")
all_gsea_activeVsCtrl_b <- all_gsea_activeVsCtrl
load("Rdata/all_gsea_activeVsDead.Rdata")
all_gsea_activeVsDead_b <- all_gsea_activeVsDead
load("Rdata/all_gsea_deadVsCtrl.Rdata")
all_gsea_deadVsCtrl_b <- all_gsea_deadVsCtrl
load("../../../Btgh_injected_morulae_SMARTSeq/src/Rmd/Rdata/all_gsea_activeVsCtrl.Rdata")
load("../../../Btgh_injected_morulae_SMARTSeq/src/Rmd/Rdata/all_gsea_activeVsDead.Rdata")
load("../../../Btgh_injected_morulae_SMARTSeq/src/Rmd/Rdata/all_gsea_deadVsCtrl.Rdata")

```

Protocol:

1. merging GSEA result for each comparison at each stage
2. choosing first 20 GOs for each stage based on abs(Normalized Enrichment Score)
3. if terms appearing in the plot for the deadVsCtrl comparison are unique to this comparison, making a second plot without the deadVsCtrl comparison (specified in plot title)

<n>

IMPORTANT: Ribosome related terms are removed from all plots and then plotted all together at the end.

# Biological Process GO terms (FIGURE for upregulated terms)

```{r GO Enrichment Test plot BP, include = TRUE, fig.height=8, fig.width=12}

# combining gsea results
all_gsea_BP_df_m <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl[[1]], showCategory = nrow(all_gsea_activeVsCtrl[[1]]@result)), fortify(all_gsea_activeVsDead[[1]], showCategory = nrow(all_gsea_activeVsDead[[1]]@result)), fortify(all_gsea_deadVsCtrl[[1]], showCategory = nrow(all_gsea_deadVsCtrl[[1]]@result))))
all_gsea_BP_df_m$comparison <- rep(c("activeVsCtrl", "activeVsDead", "deadVsCtrl"), c(nrow(all_gsea_activeVsCtrl[[1]]@result), nrow(all_gsea_activeVsDead[[1]]@result), nrow(all_gsea_deadVsCtrl[[1]]@result)))
all_gsea_BP_df_b <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl_b[[1]], showCategory = nrow(all_gsea_activeVsCtrl_b[[1]]@result)), fortify(all_gsea_activeVsDead_b[[1]], showCategory = nrow(all_gsea_activeVsDead_b[[1]]@result)), fortify(all_gsea_deadVsCtrl_b[[1]], showCategory = nrow(all_gsea_deadVsCtrl_b[[1]]@result))))
all_gsea_BP_df_b$comparison <- rep(c("activeVsCtrl", "activeVsDead", "deadVsCtrl"), c(nrow(all_gsea_activeVsCtrl_b[[1]]@result), nrow(all_gsea_activeVsDead_b[[1]]@result), nrow(all_gsea_deadVsCtrl_b[[1]]@result)))
all_gsea_BP_df_m_b <- rbind(all_gsea_BP_df_m, all_gsea_BP_df_b)
all_gsea_BP_df_m_b$stage <- rep(c("morulae", "blastocysts"), c(nrow(all_gsea_BP_df_m), nrow(all_gsea_BP_df_b)))
all_gsea_BP_df_m_b$stage <- factor(all_gsea_BP_df_m_b$stage, levels = c("morulae", "blastocysts"))
# computing absolute NES
all_gsea_BP_df_m_b$abs_NES <- abs(all_gsea_BP_df_m_b$NES)
# ordering based on abs NES
setorder(all_gsea_BP_df_m_b, stage, -abs_NES, comparison)
# taking first GOs for each stage based on NES
best_NES_terms <- all_gsea_BP_df_m_b[, .SD[1:20], by=c("stage"), .SDcols = "ID"]$ID
red_all_gsea_BP_df_m_b <- all_gsea_BP_df_m_b[ID %in% best_NES_terms]

library(DOSE)
library(tidytext)

ggplot(red_all_gsea_BP_df_m_b[!grepl("ribosom|polysome|rRNA|translation", red_all_gsea_BP_df_m_b$Description)], aes(x = GeneRatio, y = reorder_within(Description, GeneRatio, .sign), color = p.adjust)) +
      scale_color_continuous(low="red", high="blue", name = "p.adjust", guide = guide_colorbar(reverse=TRUE)) +
      geom_point(aes(size=setSize)) +
      scale_y_reordered() +
      scale_size(range=c(0.5, 5)) +
      ylab(NULL) +
      theme_dose(10) +
      facet_grid(.sign ~ stage * comparison, scales = "free_y") +
      theme(title = element_text(size = 14),
            legend.text = element_text(size = 10), legend.title = element_text(size = 8))

# now I remove the deadVsCtrl

all_gsea_BP_df_m <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl[[1]], showCategory = nrow(all_gsea_activeVsCtrl[[1]]@result)), fortify(all_gsea_activeVsDead[[1]], showCategory = nrow(all_gsea_activeVsDead[[1]]@result))))
all_gsea_BP_df_m$comparison <- rep(c("activeVsCtrl", "activeVsDead"), c(nrow(all_gsea_activeVsCtrl[[1]]@result), nrow(all_gsea_activeVsDead[[1]]@result)))
all_gsea_BP_df_b <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl_b[[1]], showCategory = nrow(all_gsea_activeVsCtrl_b[[1]]@result)), fortify(all_gsea_activeVsDead_b[[1]], showCategory = nrow(all_gsea_activeVsDead_b[[1]]@result))))
all_gsea_BP_df_b$comparison <- rep(c("activeVsCtrl", "activeVsDead"), c(nrow(all_gsea_activeVsCtrl_b[[1]]@result), nrow(all_gsea_activeVsDead_b[[1]]@result)))
all_gsea_BP_df_m_b <- rbind(all_gsea_BP_df_m, all_gsea_BP_df_b)
all_gsea_BP_df_m_b$stage <- rep(c("morulae", "blastocysts"), c(nrow(all_gsea_BP_df_m), nrow(all_gsea_BP_df_b)))
all_gsea_BP_df_m_b$stage <- factor(all_gsea_BP_df_m_b$stage, levels = c("morulae", "blastocysts"))
all_gsea_BP_df_m_b$abs_NES <- abs(all_gsea_BP_df_m_b$NES) 
setorder(all_gsea_BP_df_m_b, stage, -abs_NES, comparison)
best_NES_terms <- all_gsea_BP_df_m_b[, .SD[1:20], by=c("stage"), .SDcols = "ID"]$ID
red_all_gsea_BP_df_m_b <- all_gsea_BP_df_m_b[ID %in% best_NES_terms]

ggplot(red_all_gsea_BP_df_m_b[!grepl("ribosom|polysome|rRNA|translation|process", red_all_gsea_BP_df_m_b$Description)], aes(x = GeneRatio, y = reorder_within(Description, GeneRatio, .sign), color = p.adjust)) +
      scale_color_continuous(low="red", high="blue", name = "p.adjust", guide = guide_colorbar(reverse=TRUE)) +
      geom_point(aes(size=setSize)) +
      scale_y_reordered() +
      scale_size(range=c(0.5, 5)) +
      ylab(NULL) +
      ggtitle("Merging only activeVsCtrl and activeVsDead") +
      theme_dose(10) +
      facet_grid(.sign ~ stage * comparison, scales = "free_y") +
      theme(title = element_text(size = 14),
            legend.text = element_text(size = 10), legend.title = element_text(size = 8))

```

# Cellular Component GO terms (FIGURE for downregulated terms)

```{r GO Enrichment Test plot CC, include = TRUE, fig.height=8, fig.width=12}

all_gsea_CC_df_m <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl[[3]], showCategory = nrow(all_gsea_activeVsCtrl[[3]]@result)), fortify(all_gsea_activeVsDead[[3]], showCategory = nrow(all_gsea_activeVsDead[[3]]@result)), fortify(all_gsea_deadVsCtrl[[3]], showCategory = nrow(all_gsea_deadVsCtrl[[3]]@result))))
all_gsea_CC_df_m$comparison <- rep(c("activeVsCtrl", "activeVsDead", "deadVsCtrl"), c(nrow(all_gsea_activeVsCtrl[[3]]@result), nrow(all_gsea_activeVsDead[[3]]@result), nrow(all_gsea_deadVsCtrl[[3]]@result)))
all_gsea_CC_df_b <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl_b[[3]], showCategory = nrow(all_gsea_activeVsCtrl_b[[3]]@result)), fortify(all_gsea_activeVsDead_b[[3]], showCategory = nrow(all_gsea_activeVsDead_b[[3]]@result)), fortify(all_gsea_deadVsCtrl_b[[3]], showCategory = nrow(all_gsea_deadVsCtrl_b[[3]]@result))))
all_gsea_CC_df_b$comparison <- rep(c("activeVsCtrl", "activeVsDead", "deadVsCtrl"), c(nrow(all_gsea_activeVsCtrl_b[[3]]@result), nrow(all_gsea_activeVsDead_b[[3]]@result), nrow(all_gsea_deadVsCtrl_b[[3]]@result)))
all_gsea_CC_df_m_b <- rbind(all_gsea_CC_df_m, all_gsea_CC_df_b)
all_gsea_CC_df_m_b$stage <- rep(c("morulae", "blastocysts"), c(nrow(all_gsea_CC_df_m), nrow(all_gsea_CC_df_b)))
all_gsea_CC_df_m_b$stage <- factor(all_gsea_CC_df_m_b$stage, levels = c("morulae", "blastocysts"))
all_gsea_CC_df_m_b$abs_NES <- abs(all_gsea_CC_df_m_b$NES) 
setorder(all_gsea_CC_df_m_b, stage, -abs_NES, comparison)
best_NES_terms <- all_gsea_CC_df_m_b[, .SD[1:20], by=c("stage"), .SDcols = "ID"]$ID
red_all_gsea_CC_df_m_b <- all_gsea_CC_df_m_b[ID %in% best_NES_terms]

ggplot(red_all_gsea_CC_df_m_b[!grepl("ribosom|polysome|rRNA", red_all_gsea_CC_df_m_b$Description)], aes(x = GeneRatio, y = reorder_within(Description, GeneRatio, .sign), color = p.adjust)) +
      scale_color_continuous(low="red", high="blue", name = "p.adjust", guide = guide_colorbar(reverse=TRUE)) +
      geom_point(aes(size=setSize)) +
      scale_y_reordered() +
      scale_size(range=c(0.5, 5)) +
      ylab(NULL) +
      theme_dose(10) +
      facet_grid(.sign ~ stage * comparison, scales = "free_y") +
      theme(title = element_text(size = 14),
            legend.text = element_text(size = 10), legend.title = element_text(size = 8))

# now I remove the deadVsCtrl

all_gsea_CC_df_m <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl[[3]], showCategory = nrow(all_gsea_activeVsCtrl[[3]]@result)), fortify(all_gsea_activeVsDead[[3]], showCategory = nrow(all_gsea_activeVsDead[[3]]@result))))
all_gsea_CC_df_m$comparison <- rep(c("activeVsCtrl", "activeVsDead"), c(nrow(all_gsea_activeVsCtrl[[3]]@result), nrow(all_gsea_activeVsDead[[3]]@result)))
all_gsea_CC_df_b <- as.data.table(rbind(fortify(all_gsea_activeVsCtrl_b[[3]], showCategory = nrow(all_gsea_activeVsCtrl_b[[3]]@result)), fortify(all_gsea_activeVsDead_b[[3]], showCategory = nrow(all_gsea_activeVsDead_b[[3]]@result))))
all_gsea_CC_df_b$comparison <- rep(c("activeVsCtrl", "activeVsDead"), c(nrow(all_gsea_activeVsCtrl_b[[3]]@result), nrow(all_gsea_activeVsDead_b[[3]]@result)))
all_gsea_CC_df_m_b <- rbind(all_gsea_CC_df_m, all_gsea_CC_df_b)
all_gsea_CC_df_m_b$stage <- rep(c("morulae", "blastocysts"), c(nrow(all_gsea_CC_df_m), nrow(all_gsea_CC_df_b)))
all_gsea_CC_df_m_b$stage <- factor(all_gsea_CC_df_m_b$stage, levels = c("morulae", "blastocysts"))
all_gsea_CC_df_m_b$abs_NES <- abs(all_gsea_CC_df_m_b$NES) 
setorder(all_gsea_CC_df_m_b, stage, -abs_NES, comparison)
best_NES_terms <- all_gsea_CC_df_m_b[, .SD[1:20], by=c("stage"), .SDcols = "ID"]$ID
red_all_gsea_CC_df_m_b <- all_gsea_CC_df_m_b[ID %in% best_NES_terms]

ggplot(red_all_gsea_CC_df_m_b[!grepl("ribosom|polysome|rRNA", red_all_gsea_CC_df_m_b$Description)], aes(x = GeneRatio, y = reorder_within(Description, GeneRatio, .sign), color = p.adjust)) +
      scale_color_continuous(low="red", high="blue", name = "p.adjust", guide = guide_colorbar(reverse=TRUE)) +
      geom_point(aes(size=setSize)) +
      scale_y_reordered() +
      scale_size(range=c(0.5, 5)) +
      ylab(NULL) +
      ggtitle("Merging only activeVsCtrl and activeVsDead") +
      theme_dose(10) +
      facet_grid(.sign ~ stage * comparison, scales = "free_y") +
      theme(title = element_text(size = 14),
            legend.text = element_text(size = 10), legend.title = element_text(size = 8))

```

# Ribosome-related GO terms

```{r ribosomal GOs (FIGURE), include = TRUE}

ribo_NES_dt <- rbind(all_gsea_BP_df_m_b[grepl("ribosom|polysome|rRNA|translation", all_gsea_BP_df_m_b$Description)], all_gsea_CC_df_m_b[grepl("ribosom|polysome|rRNA|translation", all_gsea_CC_df_m_b$Description)])
ribo_NES_dt <-  ribo_NES_dt[, lapply(.SD, mean), by = c("stage", "comparison"), .SDcols = "NES"]
ggplot(ribo_NES_dt, aes(x = stage, y = NES, fill = comparison)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

```

```{r sessioninfo, echo = TRUE, results = 'markup'}

sessionInfo()

```