---
title: "Dynamics of Differentially Expressed Genes (DEGs) across preimplantation stages"
date: "January 4th, 2022"
author: 
- name: Sara Formichetti
  affiliation: EMBL Rome
  email: sara.formichetti@embl.it
output: html_document
---

```{r Setting general options, include=FALSE}

knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, echo = FALSE, warning = FALSE, message = FALSE, self.contained = TRUE)

```

```{r loading needed packages}

library("DESeq2")
library(ggplot2)
library("ggrepel")
library(data.table)

```

```{r loading necessary Rdata}

load("../../../Btgh_injected_morulae_SMARTSeq/src/Rmd/Rdata/res_all.Rdata")
load("../../../Btgh_injected_morulae_SMARTSeq/src/Rmd/Rdata/design_df.Rdata")
res_m <- res_all
design_df_m <- design_df
load("../../../Btgh_injected_blastocysts_SMARTSeq/src/Rmd/Rdata/res_all_RUVs.Rdata")
load("../../../Btgh_injected_blastocysts_SMARTSeq/src/Rmd/Rdata/design_df.Rdata")
res_b <- res_all_RUVs
design_df_b <- design_df
load("../../../Btgh_injected_2cell_SMARTSeq/src/Rmd/Rdata/res_all_RUVs.Rdata")
load("../../../Btgh_injected_2cell_SMARTSeq/src/Rmd/Rdata/design_df.Rdata")
res_2cell <- res_all_RUVs
design_df_2cell <- design_df
load("Rdata/geneID2name_w_ex.Rdata")

```


```{r reading all DEGs}

# 2-cell DEGs
DEGs_2c_activeVsDead <- fread("../../../Btgh_injected_2cell_SMARTSeq/analysis/DESeq/RUVs/all/active_Btgh_vs_dead_Btgh_DEG_by_pvalue.csv")
DEGs_2c_activeVsCtrl <- fread("../../../Btgh_injected_2cell_SMARTSeq/analysis/DESeq/RUVs/all/active_Btgh_vs_non_injected_DEG_by_pvalue.csv")
DEGs_2c_deadVsCtrl <- fread("../../../Btgh_injected_2cell_SMARTSeq/analysis/DESeq/RUVs/all/dead_Btgh_vs_non_injected_DEG_by_pvalue.csv")

# morulae DEGs
DEGs_mor_activeVsDead <- fread("../../../Btgh_injected_morulae_SMARTSeq/analysis/DESeq/active_Btgh_vs_dead_Btgh_DEG_by_pvalue.csv")
DEGs_mor_activeVsCtrl <- fread("../../../Btgh_injected_morulae_SMARTSeq/analysis/DESeq/active_Btgh_vs_non_injected_DEG_by_pvalue.csv")
DEGs_mor_deadVsCtrl <- fread("../../../Btgh_injected_morulae_SMARTSeq/analysis/DESeq/dead_Btgh_vs_non_injected_DEG_by_pvalue.csv")

# blastocysts DEGs
DEGs_blasto_activeVsDead <- fread("../../../Btgh_injected_blastocysts_SMARTSeq/analysis/DESeq/RUVs/active_Btgh_vs_dead_Btgh_DEG_by_pvalue.csv")
DEGs_blasto_activeVsCtrl <- fread("../../../Btgh_injected_blastocysts_SMARTSeq/analysis/DESeq/RUVs/active_Btgh_vs_non_injected_DEG_by_pvalue.csv")
DEGs_blasto_deadVsCtrl <- fread("../../../Btgh_injected_blastocysts_SMARTSeq/analysis/DESeq/RUVs/dead_Btgh_vs_non_injected_DEG_by_pvalue.csv")

```

# Overlap between DEGs found at the three stages

```{r venn DEGs}

library(gplots)

twocell_active <- unique(c(DEGs_2c_activeVsCtrl$gene_id[DEGs_2c_activeVsCtrl$baseMean > 10], DEGs_2c_activeVsDead$gene_id[DEGs_2c_activeVsDead$baseMean > 10]))
mor_active <- unique(c(DEGs_mor_activeVsCtrl$gene_id[DEGs_mor_activeVsCtrl$baseMean > 10], DEGs_mor_activeVsDead$gene_id[DEGs_mor_activeVsDead$baseMean > 10]))
blasto_active <- unique(c(DEGs_blasto_activeVsCtrl$gene_id[DEGs_blasto_activeVsCtrl$baseMean > 10], DEGs_blasto_activeVsDead$gene_id[DEGs_blasto_activeVsDead$baseMean > 10]))
venn(list(`2c active all` = twocell_active, `morula active all` = mor_active, `blasto active all` = blasto_active), show.plot=TRUE)

venn(list(`2c dead vs ctrl` = DEGs_2c_deadVsCtrl$gene_id[DEGs_2c_deadVsCtrl$baseMean > 10], `morula dead vs ctrl` = DEGs_mor_deadVsCtrl$gene_id[DEGs_mor_deadVsCtrl$baseMean > 10], `blasto dead vs ctrl` = DEGs_blasto_deadVsCtrl$gene_id[DEGs_blasto_deadVsCtrl$baseMean > 10]), show.plot=TRUE)

```

# DEGs expression change across all stages

```{r merging DEGs}

## rowbinding DEGs of the activeVsDead and activeVsCtrl comparisons for the three stages

comb_DEGs_dt <- rbind(DEGs_2c_activeVsDead, DEGs_2c_activeVsCtrl, DEGs_mor_activeVsDead, DEGs_mor_activeVsCtrl, DEGs_blasto_activeVsDead, DEGs_blasto_activeVsCtrl)
comb_DEGs_dt$stage <- rep(c("2c", "morulae", "blastocysts"), c((nrow(DEGs_2c_activeVsDead)+nrow(DEGs_2c_activeVsCtrl)), (nrow(DEGs_mor_activeVsDead)+nrow(DEGs_mor_activeVsCtrl)), (nrow(DEGs_blasto_activeVsDead)+nrow(DEGs_blasto_activeVsCtrl))))
comb_DEGs_dt$comparison <- rep(c("activeVsDead", "activeVsCtrl", "activeVsDead", "activeVsCtrl", "activeVsDead", "activeVsCtrl"), c(nrow(DEGs_2c_activeVsDead), nrow(DEGs_2c_activeVsCtrl), nrow(DEGs_mor_activeVsDead), nrow(DEGs_mor_activeVsCtrl), nrow(DEGs_blasto_activeVsDead), nrow(DEGs_blasto_activeVsCtrl)))
# selecting only the ones with baseMean >= 10 in that stage
comb_DEGs_dt <- comb_DEGs_dt[baseMean > 10]

## annotating list of genes with stage of diff expression 

# long to large dt
DEGs_stage <- dcast(unique(comb_DEGs_dt[, .(gene_id, gene_name, stage)]), gene_id + gene_name ~ stage, value.var = "stage")
# substituting NAs with empty string
DEGs_stage$`2c`[is.na(DEGs_stage$`2c`)] <- ""
DEGs_stage$morulae[is.na(DEGs_stage$morulae)] <- ""
DEGs_stage$blastocysts[is.na(DEGs_stage$blastocysts)] <- ""
# creating variable with all stages of diff expr
DEGs_stage$DE_stage <- gsub("\\.\\.", ".", gsub("\\.*$", "", gsub("^\\.*", "", paste(DEGs_stage$`2c`, DEGs_stage$morulae, DEGs_stage$blastocysts, sep = "."))))

```

```{r log2FC of DEGs dt}

# make list of DESeq results for the activeVsDead comparison
res_all_preimpl_activeVsDead <- list(twocell = res_2cell$active_vs_dead, morulae = res_m$active_vs_dead, blasto = res_b$active_vs_dead)

# function to select DEGs and make dt from DESeq res
get_DEGs_dt <- function (i, my_res_list, my_DEGs_dt) {
  # res as data.table
  res_dt <- as.data.table(my_res_list[[i]])
  # storing stage into column names
  names(res_dt) <- paste0(names(res_dt), ".", names(my_res_list)[[i]])
  # gene ids were the rownames of the res table
  res_dt$gene_id <- rownames(my_res_list[[i]])
  res_dt <- merge(res_dt, unique(my_DEGs_dt[, .(gene_id, gene_name)]), by = "gene_id", all = FALSE)
  return(res_dt)
}

# applying to active vs dead comparison
DEGs_res_all_preimpl_activeVsDead <- lapply(1:length(res_all_preimpl_activeVsDead), get_DEGs_dt, my_DEGs_dt = comb_DEGs_dt, my_res_list = res_all_preimpl_activeVsDead)
# merge res for all stages
DEGs_res_all_preimpl_activeVsDead_merged <- Reduce(function(...) merge(..., by = c("gene_id", "gene_name")), DEGs_res_all_preimpl_activeVsDead)
# excluding genes with baseMean < 10 at any stage
DEGs_res_all_preimpl_activeVsDead_merged <- DEGs_res_all_preimpl_activeVsDead_merged[baseMean.blasto >= 10 & baseMean.morulae >= 10 & baseMean.twocell >= 10]
# adding diff expr stage
DEGs_res_all_preimpl_activeVsDead_merged <- merge(DEGs_res_all_preimpl_activeVsDead_merged, DEGs_stage[, .(gene_id, DE_stage)], by = "gene_id")

save(DEGs_res_all_preimpl_activeVsDead_merged, file = "Rdata/DEGs_res_all_preimpl_activeVsDead_merged.Rdata") # needed for PCA with all preimpl stages using DEGs (FIGURE) in another Rmd

```

Genes coherently changing in the two stages in the plot (i.e. l2FC>=0.05 or <=-0.05 in both stages and with at least abs(l2FC)>=0.1 in one of the two stages) are colored and labeled.

```{r color and label DEGs in scatter}

## Genes coherently changing in the two stages in the plot (i.e. l2FC>=0.05 or <=-0.05 in both stages and with at least abs(l2FC)>=0.1 in one of the two stages) are colored, genes coherently changing in all three stages (i.e. l2FC>=0.05 or <=-0.05 in all three stages and with at least abs(l2FC)>=0.1 in one of the three stages) are also labeled

DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_mor <- "none"
DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_mor[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae >= 0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae >= 0.1)] <- "twocell_mor_up"
DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_mor[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae <= -0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae <= -0.1)] <- "twocell_mor_down"
DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_blasto <- "none"
DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_blasto[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto >= 0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto >= 0.1)] <- "twocell_blasto_up"
DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_blasto[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto <= -0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto <= -0.1)] <- "twocell_blasto_down"

DEGs_res_all_preimpl_activeVsDead_merged$color_all <- "none"
DEGs_res_all_preimpl_activeVsDead_merged$color_all[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae <= -0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto <= -0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.mor <= -0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto <= -0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell <= -0.1)] <- "all_down"
DEGs_res_all_preimpl_activeVsDead_merged$color_all[DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae >= 0.05 & DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto >= 0.05 & (DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.mor >= 0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto >= 0.1 | DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell >= 0.1)] <- "all_up"

DEGs_res_all_preimpl_activeVsDead_merged$label_twocell_mor <- DEGs_res_all_preimpl_activeVsDead_merged$gene_name
DEGs_res_all_preimpl_activeVsDead_merged$label_twocell_mor[DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_mor == "none"] <- ""
DEGs_res_all_preimpl_activeVsDead_merged$label_twocell_blasto <- DEGs_res_all_preimpl_activeVsDead_merged$gene_name
DEGs_res_all_preimpl_activeVsDead_merged$label_twocell_blasto[DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_blasto == "none"] <- ""
DEGs_res_all_preimpl_activeVsDead_merged$label_all <- DEGs_res_all_preimpl_activeVsDead_merged$gene_name
DEGs_res_all_preimpl_activeVsDead_merged$label_all[DEGs_res_all_preimpl_activeVsDead_merged$color_all == "none"] <- ""

write.csv(DEGs_res_all_preimpl_activeVsDead_merged, file = "../../analysis/DEGs_across_preimpl/DEGs_res_all_preimpl_activeVsDead_merged_afterFilter.csv", quote = FALSE)

```

```{r DEGs scatter, include = TRUE}

ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged[log2FoldChange.twocell < 2.5], aes(x = log2FoldChange.twocell, y = log2FoldChange.morulae, label = label_twocell_mor)) +
  geom_point(aes(color = color_twocell_mor), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 30) +
  scale_colour_manual(values = c("twocell_mor_up" = "red", "twocell_mor_down" = "blue", "none" = "grey"))
ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged[log2FoldChange.twocell < 2.5], aes(x = log2FoldChange.twocell, y = log2FoldChange.blasto, label = label_twocell_blasto)) +
  geom_point(aes(color = color_twocell_blasto), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 30) +
  scale_colour_manual(values = c("twocell_blasto_up" = "red", "twocell_blasto_down" = "blue", "none" = "grey"))

ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged[log2FoldChange.twocell < 2.5], aes(x = log2FoldChange.twocell, y = log2FoldChange.morulae, label = label_twocell_mor)) +
  geom_point(aes(color = color_twocell_mor), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 30) +
  ggtitle("with axes limits") +
  xlim(c(-0.5,0.5)) +
  ylim(c(-0.5,0.5)) +
  coord_fixed(ratio = 1) +
  scale_colour_manual(values = c("twocell_mor_up" = "red", "twocell_mor_down" = "blue", "none" = "grey"))
ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged[log2FoldChange.twocell < 2.5], aes(x = log2FoldChange.twocell, y = log2FoldChange.blasto, label = label_twocell_blasto)) +
  geom_point(aes(color = color_twocell_blasto), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 30) +
  ggtitle("with axes limits") +
  xlim(c(-0.5,0.5)) +
  ylim(c(-0.5,0.5)) +
  coord_fixed(ratio = 1) +
  scale_colour_manual(values = c("twocell_blasto_up" = "red", "twocell_blasto_down" = "blue", "none" = "grey"))

```

## FIGURE

Genes coherently changing in the two stages in the plot (i.e. l2FC>=0.05 or <=-0.05 in both stages and with at least abs(l2FC)>=0.1 in one of the two stages) are colored, genes coherently changing **in all three stages** (i.e. l2FC>=0.05 or <=-0.05 in all three stages and with at least abs(l2FC)>=0.1 in one of the three stages) are also labeled.

```{r DEGs scatter w axis limits (FIGURE), include = TRUE}

ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged, aes(x = log2FoldChange.twocell, y = log2FoldChange.morulae, label = label_all)) +
  geom_point(aes(color = color_twocell_mor), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 100) +
  scale_colour_manual(values = c("twocell_mor_up" = "red", "twocell_mor_down" = "blue", "none" = "grey")) +
  xlim(c(-0.5,0.5)) +
  ylim(c(-0.5,0.5)) +
  coord_fixed(ratio = 1) +
  theme(legend.position = "none")
ggplot(data = DEGs_res_all_preimpl_activeVsDead_merged, aes(x = log2FoldChange.twocell, y = log2FoldChange.blasto, label = label_all)) +
  geom_point(aes(color = color_twocell_blasto), alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 100) +
  scale_colour_manual(values = c("twocell_blasto_up" = "red", "twocell_blasto_down" = "blue", "none" = "grey")) +
  xlim(c(-0.5,0.5)) +
  ylim(c(-0.5,0.5)) +
  coord_fixed(ratio = 1) +
  theme(legend.position = "none")

```

## Correlations - all merged DEGs

```{r compute correlations, echo = TRUE}

cor(DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell, DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.morulae, method = "pearson")
cor(DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.twocell, DEGs_res_all_preimpl_activeVsDead_merged$log2FoldChange.blasto, method = "pearson")

```

## Correlations - only coherently changing DEGs

```{r compute correlations 2, echo = TRUE}

coherent_DEGs_twocell_mor_dt <- DEGs_res_all_preimpl_activeVsDead_merged[DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_mor != "none"]
coherent_DEGs_twocell_blasto_dt <- DEGs_res_all_preimpl_activeVsDead_merged[DEGs_res_all_preimpl_activeVsDead_merged$color_twocell_blasto != "none"]

cor(coherent_DEGs_twocell_mor_dt$log2FoldChange.twocell, coherent_DEGs_twocell_mor_dt$log2FoldChange.morulae, method = "pearson")
cor(coherent_DEGs_twocell_blasto_dt$log2FoldChange.twocell, coherent_DEGs_twocell_blasto_dt$log2FoldChange.blasto, method = "pearson")

```

```{r sessioninfo, echo = TRUE, results = 'markup'}

sessionInfo()

```